\bf Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation

### 2025-02-10 paper/main.tex
Updates from Overleaf

@@

\if1\blind
{
  \title{\bf Longitudinal Principal Manifold Estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Longitudinal Principal Manifold Estimation}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation. %\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
\end{abstract}

@@

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

@@

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.

@@

Hereafter, we utilize the following notation to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each image $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension
\begin{align}\label{eq: F form}
    \begin{aligned}
        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R},\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
    \end{aligned}
\end{align}
We define a longitudinal principal manifold as follows.
\begin{definition}
  \label{def:lpme} Given a collection $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$ of data clouds observed at a series of time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ as follows
\begin{align}\label{eq:newKappa}
      \notag\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
\end{align}
\sloppy where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation \eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u: \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

@@

\begin{figure}
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D3}}}
  \caption{{\footnotesize Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the polar coordinates results in a manifold that is not self-intersecting.}}
  \label{fig:unit_circle_augmentation}
\end{figure}

Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space. This approach to data augmentation offers an alternative to manual partitioning used by \cite{mengPrincipalManifoldEstimation2021}. Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate. This source of error is avoided using our proposed data augmentation approach. Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold. This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances. However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.

@@

To assess the performance of the LPME algorithm (Algorithm \ref{alg:lpme}), we use simulation studies to compare how closely (in terms of reduction of MSE) the LPME-estimated underlying functions approximate the true data generating functions in comparison with that of three alternative approaches. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, as well as $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, we consider differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time. The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}. Based on our observations of systematic noise sources in imaging data, we add various types of terms to these functions at each visit. In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between visits and represent structural changes in the underlying manifold over time. The function $g(\cdot)$ may show change with respect to time that is constant (no change), linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier. The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between visits. Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean 1 and variance that is varied between visits. These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.

\begin{table}[ht]
  \centering
  \resizebox{\textwidth}{!}{\begin{tabular}{|c c c c c|}
    \hline
    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
    \hline
    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
    \hline
  \end{tabular}}
  \caption{{\footnotesize Embedding functions used for simulation studies. Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change. Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}}
  \label{table:simulation_embeddings}
\end{table}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal. The sample size for each time point is set to 1,000 observations. Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 7,776 times. \if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi

@@

\begin{figure}
  \centering
  \includegraphics[height=9cm]{sim_case1} 
  \caption{{\footnotesize Simulation Case 1. Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).}}
  \label{fig:sim_case1}
\end{figure}

@@

@@

As discussed previously, meaningful between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images. This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates. To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of AD.

>>>>

\newcommand{\revision}[1]{{\color{blue}}}

@@

\if1\blind
{
  \title{\bf Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Longitudinal structural magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampi and thalami using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation. %\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
\end{abstract}

@@

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. Similarly, \cite{akudjedu2018comparative} found that FIRST and FreeSurfer showed strong correlation with manual segmentation for the caudate, but relatively poor absolute agreement and moderate relative consistency when applied to the hippocampus. When comparing estimated subcortical volumes obtained using FIRST and FreeSurfer in two consecutive yearly study visits, \cite{quilis-sancho2020comparative} found that considerable variability was present in the relative differences in subcortical volume estimates, even after removing outliers from analysis. Furthermore, this variability was much greater for FIRST than FreeSurfer, with standard deviations for relative volumetric difference in the left and right hippocampus reaching 12.2\% and 11.3\%, respectively \citep{quilis-sancho2020comparative}. Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

@@

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach assumes that the progression trajectory for each subject must be parallel to a straight reference trajectory in some latent space, which may be overly restrictive in the case of heterogeneous conditions like AD. Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.

@@

Hereafter, we utilize the following notation to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each image $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension
\begin{align}\label{eq: F form}
    \begin{aligned}
        F:\ \ & \mathbb{R}\times\mathbb{R}^d \rightarrow \mathbb{R}\times\mathbb{R}^D,\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
    \end{aligned}
\end{align}
We define a longitudinal principal manifold as follows.
\begin{definition}
  \label{def:lpme} Given a collection $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$ of data clouds observed at a series of time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ as follows
\begin{align}\label{eq:newKappa}
      \notag\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
\end{align}
\sloppy where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation \eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u: \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

@@

\begin{figure}
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D3}}}
  \caption{{\footnotesize Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the polar coordinates results in a manifold that is not self-intersecting.}}
  \label{fig:unit_circle_augmentation}
\end{figure}
Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space. This approach to data augmentation offers an alternative to manual partitioning used by \cite{mengPrincipalManifoldEstimation2021}. Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate. This source of error is avoided using our proposed data augmentation approach. Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold. This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances. However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.

@@

To assess the performance of the LPME algorithm (Algorithm \ref{alg:lpme}), we use simulation studies to compare how closely (in terms of reduction of MSE) the LPME-estimated underlying functions approximate the true data generating functions in comparison with that of three alternative approaches. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, as well as $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, we consider differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time. The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}. We incorporate several sources of systematic variability to reflect the patterns of noise commonly observed in imaging data. In these functions, we specify $\alpha = 1 - \zeta\mathbf{g}(t) + \upsilon_{\text{a}}$ and $\beta = 1 + \upsilon_{\text{p}}$, where $\zeta$ and the functional form of $g(\cdot)$ are varied between visits to represent structural changes in the underlying manifold over time. The function $g(\cdot)$ may show change with respect to time that is constant ($g(t) = 0$), linear ($g(t) = t$), or quadratic ($g(t) = t^2$) with $\zeta$ serving as a scalar multiplier. The values $\upsilon_{\text{a}}$ and $\upsilon_{\text{p}}$, denoting random fluctuations in the manifold amplitude and period between visits, are randomly sampled from a normal distribution with mean zero and variance that is varied between simulation runs. Finally, the $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between simulation runs.

Of particular interest is simulation case 8, in which the simulated data takes the form of a sphere with $D = 3$. Because $t \geq 0$ for all simulated visits, the linear and quadratic forms of $g(\cdot)$ result in a sphere with a decreasing radius, approximating atrophy in a subcortical structure.

\begin{table}[ht]
  \centering
  \resizebox{\textwidth}{!}{\begin{tabular}{|c c c c c|}
    \hline
    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
    \hline
    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \iota$ & $-3 \leq r \leq 3$ \\
    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \iota$ & $-1 \leq r \leq 1$ \\
    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \iota$ & $0 \leq r \leq 3\pi$ \\
    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
    \hline
  \end{tabular}}
  \caption{{\footnotesize Embedding functions used for simulation studies. Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change. Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}}
  \label{table:simulation_embeddings}
\end{table}

%\begin{table}[ht]
%  \centering
%  \resizebox{\textwidth}{!}{\begin{tabular}{|c c c c c|}
%    \hline
%    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
%    \hline
%    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
%    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
%    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
%    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
%    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
%    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
%    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
%    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
%    \hline
%  \end{tabular}}
%  \caption{{\footnotesize Embedding functions used for simulation studies. Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change. Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}}
%  \label{table:simulation_embeddings}
%\end{table}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\iota, \upsilon_{\text{a}}, \upsilon_{\text{p}} \in \left\{0.05, 0.1, 0.25, 0.5\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic; 5) $\zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5\right\}$. The sample size for each time point is set to 1,000 observations. Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 5,760 times. \if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme}{\texttt{https://github.com/rjzielinski/lpme}}.}\fi Note that while discrete time points were used in these simulations, the spline model is also able to accommodate continuous time values, as seen in the ADNI application given in Section \ref{s:application}.

@@

\begin{figure}
  \centering
  \includegraphics[height=9cm]{sim_case1} 
  \caption{{\footnotesize Simulation Case 1. Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles). Noise introduced in the data generation process causes the observed data to deviate from the true manifold. The PME and principal curve estimates tend to closely follow the data at each time point, while the LPME estimates remain closer to the true manifold.}}
  \label{fig:sim_case1}
\end{figure}

@@

INCLUDE VOLUME ESTIMATES FROM CASE 8 TO INDICATE PERFORMANCE IN ATROPHY SITUATION

@@

As discussed previously, substantial between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images. This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates. To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of AD.
